---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
--- 
### Loading and preprocessing the data

```{r setoptions, echo=TRUE, comment=NA}
file<-"activity.csv"
data<-read.csv(file, header = T, sep = ",")
```  
    
      
### What is mean total number of steps taken per day?

```{r}
data$date<-as.Date(data$date)
stepsdate<-dplyr::group_by(data, date)
stepsdaily<-dplyr::summarize(stepsdate, stepsdaily = sum(steps, na.rm=TRUE))
hist(stepsdaily$stepsdaily)
summary(stepsdaily$stepsdaily)

# I prefer either to exclude the days with 0 values or to not remove NA to get more clear understanding
stepsdaily<- stepsdaily[stepsdaily$stepsdaily!=0,]
hist(stepsdaily$stepsdaily)
summary(stepsdaily$stepsdaily)
```
##### Mean of the total number of steps taken per day equals 9354 or 1076* 
##### Median of the total number of steps taken per day equals 10395 or 10765* 
##### (*) - excluding NA days

### What is the average daily activity pattern?

```{r}
stepsinterval<-dplyr::group_by(data, interval)
stepsavg<-dplyr::summarize(stepsinterval, stepsavg = mean(steps, na.rm=TRUE))
with(stepsavg, plot(interval, stepsavg, type="l"))

stepsavg$stepsavg<-round(stepsavg$stepsavg,2)
stepsavg[stepsavg$stepsavg==max(stepsavg$stepsavg),]
```
##### Averaged across all the days days 5-minute interval number 835 (8:35 am) 
##### contains the maximum number of steps   

### Imputing missing data

```{r}
colSums(is.na(data))
``` 
##### Total missing values in the dataset equal 2304 
##### There are 8 whole days with no step data. Thus I suggest to impute the missing values based on averaged intervals across 
##### the set.

```{r}
stepsavg$stepsavg<-as.integer(round(stepsavg$stepsavg))
data[is.na(data$steps), 1] <- rep(stepsavg$stepsavg, 2304/288)
head(data)

stepsdate<-dplyr::group_by(data, date)
stepsdaily<-dplyr::summarize(stepsdate, stepsdaily = sum(steps))
hist(stepsdaily$stepsdaily)

summary(stepsdaily$stepsdaily)
```

##### Mean of the total number of steps taken per day equals 10766
##### Median of the total number of steps taken per day equals 10762
###### Inference: data imputing significantly decreased number of days with 0-5000 step count and noticeably augmented number of days  
###### with 10000-15000 step count. There is a 3041 step augmentation in 1st quartile, mean and median values increased by   
###### 1412 and 367 steps respectively.  
###### While ignoring NA at the beginning of EDA does't give such a large differene. Due to artificially aumented days of mesurments, 1st and 3rd quartiles 
###### increased, while mean and median values remaind almost the same.


### Are there differences in activity patterns between weekdays and weekends?

```{r message=FALSE}
# Preparing data
data$weekday<-weekdays(data$date)
weekday<-unique(data$weekday)[1:5]
weekend<-unique(data$weekday)[6:7]

data$weekdayORweekend[data$weekday %in% weekday]<-"weekday"
data$weekdayORweekend[data$weekday %in% weekend]<-"weekend"
data$weekdayORweekend<-factor(data$weekdayORweekend)

stepsinterval<-dplyr::group_by(data, interval, weekdayORweekend)
stepsavg<-dplyr::summarize(stepsinterval, stepsavg = mean(steps))
```

```{r, comment=NULL}
# Plotting
library(ggplot2)
ggplot(stepsavg, aes(interval, stepsavg, colour=weekdayORweekend)) +
geom_line() + facet_wrap(~ weekdayORweekend,nrow=2)

# Optional
library(lattice)
xyplot(stepsavg~interval|factor(weekdayORweekend),
       data=stepsavg,
       type = "l",
       xlab = "Interval",
       ylab = "Number of steps",
       layout=c(1,2))
```



